Reducing Linearizability of Priority Queues and More Data Structures to State Reachability

### 2017-04-21 appendix-EPQ.tex
modify

The notion of left-right constraint used for extended priority queue is inspired by left-right constraint of queue \cite{Bouajjani:2015}. Let us introduce the notion of left-right constraint, which is a graph, and the existence of cycle though item with maximal priority in it correspond to the existence of the case in \figurename~\ref{fig:introduce gap for EPQ1Lar}.

\begin{definition}\label{def:left-right constraint for matched put and rm operations}
Given a data-differentiated $\textit{pri}_x$-execution $e$ without $\textit{rm}(\textit{empty})$. Let $\textit{put}(x,\textit{pri}_x)$ and $\textit{rm}(x)$ be the only  method events of $e$ with priority $\textit{pri}_x$. The left-right constraint of $\textit{put}(x,\textit{pri}_x)$ and $\textit{rm}(x)$ is the graph $G$ where:

\begin{itemize}
\setlength{\itemsep}{0.5pt}
\item[-] the nodes are the items of $e$, to which we add a node,

\item[-] there is an edge from item $d_1$ to $x$, if $\textit{put}(d_1,\_)$ happens before $\textit{put}(x,\textit{pri}_x)$ or $\textit{rm}(x)$,

\item[-] there is an edge from $x$ to item $d_1$, if $\textit{rm}(x)$ happens before $\textit{rm}(d_1)$ or $\textit{rm}(d_1)$ does not exists in $h$,

When there is a cycle $d_1 \rightarrow \ldots \rightarrow d_m \rightarrow x \rightarrow d_1$ through item with maximal priority (for example, $x$) in $G$, we say that $x$ is covered by $d_1,\ldots,d_m$. To state the effectiveness of left-right constraint, take the execution in \figurename~\ref{fig:introduce gap for EPQ1Lar} as an example, we can see that $b$ is covered by $f,e,d$. We need to prove that getting rid of cycle though item with maximal priority in left-right constraint is enough for ensure linearizable w.r.t $\textit{MS}(\seqPQ_1^{>})$, as stated by the following lemma:

@@

Given a history $h$, operation sets $S_1$, $S_2$ and sequences $l_1$ and $l_2$. Let $h_1 = h \vert{S_1}$ and $h_2 = h \vert{S_2}$. Assume that $h_1 \sqsubseteq l_1$, $h_2 \sqsubseteq l_2$, and $S_1 \cup S_2$ contains all operations of $h$. Then, there exists a sequence $l$, such that $h \sqsubseteq l$, $l \vert{S_1} = l_1$ and $l \vert{S_2} = l_2$.
\end{restatable}

@@

Given a history $h$ and a operation $o \in S_2$, let $\textit{MB}(o) = \{ o' \vert o' \in S_1$ and $o'$ happens before $o$ in $h \}$, let $\textit{SBI}(o) = \textit{min}\{ i \vert l_1[0,i]$ contains all elements of $\textit{MB}(o) \}$.

@@

It is obvious that $l \vert{S_1} = l_1$ and $l \vert{S_2} = l_2$, and it remains to prove that $h \sqsubseteq l$. We prove this by contradiction. Assume that $h$ is not linearizable with respect to $l$. Then there must be two operations of $h$, such that $o_1 <_{hb} o_2$ in $h$ but $o_2$ before $0_1$ in $l$. Since $l \vert{S_1} = l_1$, $l \vert{S_2} = l_2$, and $h_1 \sqsubseteq l_1$, $h_2 \sqsubseteq l_2$, it is easy to see that it is impossible that $o_1,o_2 \in S_1$ or $o_1,o_2 \in S_2$. There are only two possibilities:

@@

We deal with the case of $R = \seqPQ_1^{>}$ , and other cases can be similarly dealt with.

To prove the $\textit{only if}$ direction, given $e \sqsubseteq \textit{MS}(\seqPQ_1^{>})$ and such $\textit{pri}$ and $x$. Since $e \sqsubseteq \textit{MS}(\seqPQ_1^{>})$ with witness $x$, we know that $e \sqsubseteq u \cdot \textit{put}(x,\textit{pri}) \cdot v \cdot \textit{rm}(x) \cdot w$, where $u$, $v$, $w$, $x$ and $\textit{pri}$ satisfy the guard of $\seqPQ_1^{>}$. Let $u'$, $v'$ and $w'$ be obtained from $u$, $v$ and $w$ by erasing all items with priority incomparable with $\textit{pri}$, respectively. It is not hard to see that $u'$, $v'$, $w'$, $x$ and $\textit{pri}$ satisfy the guard of $\seqPQ_1^{>}$, and then $e \sqsubseteq l = u' \cdot \textit{put}(x,\textit{pri}) \cdot v' \cdot \textit{rm}(x) \cdot w' \in \textit{MS}(\seqPQ_1^{>})$.

To prove the $\textit{if}$ direction, given $e' = \textit{pri-Exec}(e)$ and such $x$ and $\textit{pri}$. Since $e' \sqsubseteq \textit{MS}(\seqPQ_1^{>})$ with witness $x$, we know that $e' \sqsubseteq l_1 = u \cdot \textit{put}(x,\textit{pri}) \cdot v \cdot \textit{rm}(x) \cdot w$, where $u$, $v$, $w$, $x$ and $\textit{pri}$ satisfy the guard of $\seqPQ_1^{>}$. Let $O_c$ be the set of operations in $e$ that have priorities comparable with $\textit{pri}$, and Let $O_i$ be the set of operations in $e$ that have priorities incomparable with $\textit{pri}$. It is obvious that $l_1$ is the linearization of $e \vert_{O_c}$. By Lemma \ref{lemma:merge two linearization}, there exists sequence $l$, such that $e \sqsubseteq l$, and $l \vert_{O_c} = l_1$. Then $l = u' \cdot \textit{put}(x,\textit{pri}) \cdot v' \cdot \textit{rm}(x) \cdot w'$, where $u' \vert_{O_c} = u$, $v' \vert_{O_c} = v$ and $w' \vert_{O_c} = w$. Since $\textit{pri}$ is one of maximal priorities in $e$, and the predicates of guards of $\seqPQ_1^{>}$ does not restrict $O_i$, it is easy to see that $l \in \textit{MS}(\seqPQ_1^{>})$ and then $e \sqsubseteq l$ with witness $x$. \qed
\end {proof}

We can see that $\textit{UVSet}_i(e,x) \cap \textit{UVSet}_j(e,x) = \emptyset$ for any $i \neq j$.

@@

\LinEqualsConstraintforEPQOneLar*

>>>>

\subsection{Proof of Lemma \ref{lemma:pri execution is enough}}
\label{sec:appendix proof of Lemma pri execution is enough}

@@

Given an execution $e$, operation sets $S_1$, $S_2$ and sequences $l_1$ and $l_2$. Let $e_1 = e \vert{S_1}$ and $e_2 = e \vert{S_2}$. Assume that $e_1 \sqsubseteq l_1$, $e_2 \sqsubseteq l_2$, and $S_1 \cup S_2$ contains all operations of $e$. Then, there exists a sequence $l$, such that $e \sqsubseteq l$, $l \vert{S_1} = l_1$ and $l \vert{S_2} = l_2$.
\end{restatable}

@@

Given an execution $e$ and a operation $o \in S_2$, let $\textit{MB}(o) = \{ o' \vert o' \in S_1$ and $o'$ happens before $o$ in $h \}$, let $\textit{SBI}(o) = \textit{min}\{ i \vert l_1[0,i]$ contains all elements of $\textit{MB}(o) \}$.

@@

It is obvious that $l \vert{S_1} = l_1$ and $l \vert{S_2} = l_2$, and it remains to prove that $e \sqsubseteq l$. We prove this by contradiction. Assume that $e \not \sqsubseteq l$. Then there must be two operations $o_1$, $o_2$ of $e$, such that $o_1 <_{hb} o_2$ in $e$ but $o_2$ before $0_1$ in $l$. Since $l \vert{S_1} = l_1$, $l \vert{S_2} = l_2$, and $h_1 \sqsubseteq l_1$, $h_2 \sqsubseteq l_2$, it is easy to see that it is impossible that $o_1,o_2 \in S_1$ or $o_1,o_2 \in S_2$. There are only two possibilities:

@@

We deal with the case of $\Gamma = \mathsf{MatchedMaxPriority}^{>}$ , and other cases can be similarly dealt with.

To prove the $\textit{only if}$ direction, assume that $e \sqsubseteq l$, $\mathsf{MatchedMaxPriority\text{-}Seq}(l,x)$ holds and $p$ is the priority of $x$. We can see that $e \sqsubseteq u \cdot \textit{put}(x,p) \cdot v \cdot \textit{rm}(x) \cdot w$. Let $u'$, $v'$ and $w'$ be obtained from $u$, $v$ and $w$ by erasing all items with priority incomparable with $p$, respectively. Since the predicates in $\mathsf{MatchedMaxPriority\text{-}Seq}(l,x)$ does not restrict the values of priorities incomparable with $p$, is not hard to see that $e \sqsubseteq l'$ and $\mathsf{MatchedMaxPriority\text{-}Seq}(l',x)$ holds, where $l' = u' \cdot \textit{put}(x,\textit{pri}) \cdot v' \cdot \textit{rm}(x) \cdot w'$.

To prove the $\textit{if}$ direction, given $e' = e \vert_{\prec p}$. By assumption, $e' \sqsubseteq l$, $\mathsf{MatchedMaxPriority\text{-}Seq}(l,x)$ holds and $p$ is the priority of $x$, and then we can see that $e' \sqsubseteq l_1 = u \cdot \textit{put}(x,p) \cdot v \cdot \textit{rm}(x) \cdot w$. Let $O_c$ be the set of operations in $e$ that have priorities comparable with $p$, and Let $O_i$ be the set of operations in $e$ that have priorities incomparable with $p$. It is obvious that $l_1$ is the linearization of $e \vert_{O_c}$. By Lemma \ref{lemma:merge two linearization}, there exists sequence $l$, such that $e \sqsubseteq l$, and $l \vert_{O_c} = l_1$. Then $l = u' \cdot \textit{put}(x,p) \cdot v' \cdot \textit{rm}(x) \cdot w'$, where $u' \vert_{O_c} = u$, $v' \vert_{O_c} = v$ and $w' \vert_{O_c} = w$. Since $p$ is one of maximal priorities in $e$, and the predicates in $\mathsf{MatchedMaxPriority\text{-}Seq}(l,x)$ does not restrict the values with priority $O_i$, it is easy to see that $\mathsf{MatchedMaxPriority\text{-}Seq}(l,x)$ holds. \qed
\end {proof}

Let us introduce $\textit{UVSet}(e,x)$, which intuitively contains all pairs of operations that should be putted before $\textit{rm}(x)$ when construction linearization of $x$ according to $\mathsf{MatchedMaxPriority}$. Let $\textit{UVSet}_1(e$, $x) = o \vert$ either $o <{\textit{hb}} \textit{put}(x,\_)$ or $\textit{rm}(x)$, or there exists $o'$ with the same value of $o$, such that $o' <{\textit{hb}} \textit{put}(x)$ or $\textit{rm}(x) \}$. For each $i > 1$, let $\textit{UVSet}_{\textit{i+1}}(e,x) = o \vert$ $o \notin \textit{UVSet}_k$ for each $k \leq i$, and either $o$ happens before some operation $o' \in \textit{UVSet}_i(e,x)$, or there exists $o''$ with the same value of $o$, and $o''$ happens before some operation $o' \in \textit{UVSet}_i(e,x) \}$. Let $\textit{UVSet}(e,x) = \textit{UVSet}_1(e,x) \cup \ldots$. Note that it is possible that $\textit{UVSet}_i(e,x) \cap \textit{UVSet}_j(e,x) = \emptyset$ for any $i \neq j$.

@@
